---
title: "Introduction to Shiny"
subtitle: "Session 5: *Shiny modules, working with data*"
author: "Danny Smith"
format:
  revealjs:
    theme: dark
    width: 1280
    height: 720
    auto-animate: true
    slide-number: true
    chalkboard: false
    preview-links: auto
    footer: "**Introduction to Shiny** Session 5: *Shiny modules, working with data*"
    self-contained: true
    reference-location: document
---
  
# Housekeeping

```{r}
#| output: false
library(shiny)
```

## Course outline
  
1. Introduction to Shiny
2. Shiny UI - layout, styling and theming
3. Shiny UI - dynamic UI and extensions
4. Reactivity and the user session
5. *Shiny modules, working with data*
6. Application structure and workflow

## Recap

&nbsp;

### Last Session

4. *Reactivity and the user session*

&nbsp;

::: {.incremental}
* Homework
* Readings
* Questions?
:::

## Resources

::: {.fragment}
#### Session materials

[GitHub repo](https://github.com/gorcha/r_shiny_intro_data)
:::

&nbsp;

::: {.fragment}
#### Online resources

[RStudio: Databases using R](https://solutions.rstudio.com/db/)
:::

## This session

::: {.incremental}
* Intro to Shiny modules
* Modules in practice
* Data for Shiny applications
* Working with databases
* Working with user data
:::

# Intro to Shiny modules

&nbsp;

📝 session_05/module_counter.R

## What are shiny modules?

::::: {.columns}

:::: {.column width="50%"}
Shiny equivalent of functions!

::: {.incremental}
* Compartmentalise code
  - Group UI components with a single return value
  - Self-contained - great for testing!

* Reuse complex sets of components
  - Simplify app structure
:::
::::

:::: {.column width="50%"}
::: {.fragment}
Module structure
:::

::: {.incremental}
* A module has a UI and server function
  - Define UI components
  - Define behaviour
  - Configurable using arguments
  - Include in app UI and server
  - Linked by ID
:::
::::

:::::

## Module UI

::::: {.columns}

:::: {.column width="40%"}
::: {.incremental}
* Function taking the module ID as input
* Other arguments to configure UI
* Return UI object
* IDs use *namespacing*
  - Avoid ID conflicts
  - Use the `NS()` function
:::
::::

:::: {.column width="60%"}
```{.r code-line-numbers="|1|5|4-7|2|5,6"}
counterUI <- function(id, label = "Counter") {
  ns <- NS(id)
  
  tagList(
    actionButton(ns("button"), label = label),
    verbatimTextOutput(ns("out"))
  )
}
```
::::

:::::

## Namespacing

Namespacing allows us to avoid conflicts between component IDs in the module and IDs used elsewhere

::::: {.columns}

:::: {.column width="50%"}
::: {.incremental}
* Creates an ID combining the module and component ID
  - Component ID is known when creating the module
  - Module ID is defined when the module is called
* Only necessary when creating UI
:::
::::

:::: {.column width="50%"}
::: {.incremental}
* The `NS()` function combines the module ID and component IDs
  ```{r}
  #| echo: true
  #| code-line-numbers: false
  NS("module_id", "component_id")
  ```
* `ns <- NS(id)` creates a helper function for the current module
  ```{r}
  #| echo: true
  #| code-line-numbers: false
  ns <- NS("module_id")
  ns("component_id")
  ```
:::
::::

:::::

## Module server

::::: {.columns}

:::: {.column width="40%"}
::: {.incremental}
* Function taking the module ID as input
* Other arguments to configure server behaviour
* Return `moduleServer()`
  * Module ID
  * Server function
* Namespacing in the server function is automatic
:::
::::

:::: {.column width="60%"}
```{.r code-line-numbers="|1|12|2-17|3|4-16|7,11"}
counterServer <- function(id, prefix = NULL) {
  moduleServer(
    id,
    function(input, output, session) {
      count <- reactiveVal(0)
      
      observeEvent(input$button, {
        count(count() + 1)
      })
      
      output$out <- renderText({
        paste0(prefix, count())
      })
      
      count
    }
  )
}
```
::::

:::::

## Modules in an application

::::: {.columns}

:::: {.column width="40%"}
::: {.incremental}
* Call the module UI function in your app UI
* Call the module server function in your app server function
* Modules can be used multiple times with different IDs
:::
::::

:::: {.column width="60%"}
```{.r code-line-numbers="|8|14|9,16"}
library(shiny)

source("module_counter.R", local = TRUE)

ui <- fixedPage(
  titlePanel("Counters"),
  fixedRow(
    column(6, counterUI("counter")),
    column(6, counterUI("counter2", "My 2nd counter"))
  )
)

server <- function(input, output, session) {
  counterServer("counter")
  
  counterServer("counter2")
}

shinyApp(ui, server)
```
::::

:::::

::: aside
📝 session_05/module_counter_app.R
:::

## App function

::: {.notes}
Source module_counter.R

Run `counterApp()`
:::

Good practice to create an *app function* for module testing

::::: {.columns}

:::: {.column width="40%"}
::: {.incremental}
* Function wraps a minimal application demonstrating the module
* Return a `shinyApp()` object
* Call from the console
  - Allow specification of module args
:::
::::

:::: {.column width="60%"}
```{.r code-line-numbers="|2-8|10|1"}
counterApp <- function(label = "Counter", prefix = "The current count is: ") {
  ui <- fluidPage(
    counterUI("counter", label),
  )
  
  server <- function(input, output, session) {
    counterServer("counter", prefix)
  }
  
  shinyApp(ui, server)
}
```

&nbsp;

::: {.fragment}
```{.r code-line-numbers="false"}
counterApp(
  "My button",
  "My button has been pressed this many times: "
)
```
:::
::::

:::::

## Documentation

::::: {.columns}

:::: {.column width="40%"}
📦 [roxygen2](https://roxygen2.r-lib.org/)

::: {.incremental}
* Document module functions with roxygen tags
  - Same conventions as documenting a function
* App function is a form of documentation
  - Equivalent of examples
:::
::::

:::: {.column width="60%"}
```{.r filename="module_counter.R" code-line-numbers="|1-6|16-21|41-59"}
#' Counter module UI
#'
#' @param id Module ID.
#' @param label Label for the counter button.
#'
#' @return A tagList containing counter module UI elements.
counterUI <- function(id, label = "Counter") {
  ns <- NS(id)
  
  tagList(
    actionButton(ns("button"), label = label),
    verbatimTextOutput(ns("out"))
  )
}

#' Counter module server
#'
#' @param id Module ID.
#' @param prefix Prefix for counter value text output.
#'
#' @return The counter as a reactive value.
counterServer <- function(id, prefix = NULL) {
  moduleServer(
    id,
    function(input, output, session) {
      count <- reactiveVal(0)
      
      observeEvent(input$button, {
        count(count() + 1)
      })
      
      output$out <- renderText({
        paste0(prefix, count())
      })
      
      count
    }
  )
}

#' Counter module application
#'
#' Create a simple example application using the counter module.
#' 
#' @param label Label for the counter button.
#' @param prefix Prefix for counter value text output.
#'
#' @return A shinyApp object
counterApp <- function(label = "Counter", prefix = "The current count is: ") {
  ui <- fluidPage(
    counterUI("counter", label),
  )
  
  server <- function(input, output, session) {
    counterServer("counter", prefix)
  }
  
  shinyApp(ui, server)
}
```
::::

:::::

# Modules in practice

## Reactive modules

Modules often need to be integrated with the reactive graph of your application

* Return values to use in your server function
* Respond to changes in inputs elsewhere in your application

To achieve this we can pass *reactive* inputs and outputs to and from the module server

## Reactive server outputs

::::: {.columns}

:::: {.column width="40%"}
::: {.incremental}
* Return a value from the module server function
* Must be reactive object
  - `reactive()`
  - `reactiveValues()`
* List for multiple values
* Assign result of module in server function
:::
::::

:::: {.column width="60%"}
```{.r code-line-numbers="|4-13|9-12|5-7,10|11|17-26|18|21,25"}
selectDatasetServer <- function(id) {
  moduleServer(
    id,
    function(input, output, session) {
      dataset <- reactive({
        dataset <- as.data.frame(get(input$dataset, "package:datasets"))
      })
      
      list(
        data = dataset,
        name = reactive(input$dataset)
      )
    }
  )
}

server <- function(input, output, session) {
  dataset <- selectDatasetServer("dataset_select")
  
  output$module_out <- renderText({
    paste0("Selected dataset `", dataset$name(), "`")
  })
  
  output$table <- renderTable({
    dataset$data()
  })
}
```
::::

:::::

::: aside
📝 session_05/module_reactive.R
:::

## Reactive server inputs

::::: {.columns}

:::: {.column width="40%"}
::: {.incremental}
* Use a reactive value from elsewhere in the application
* Pass as argument to module server function
  - The reactive object, not the value
* Use like a normal reactive
:::
::::

:::: {.column width="60%"}
```{.r code-line-numbers="|1|9|21-24|19,23"}
selectVariableServer <- function(id, dataset) {
  moduleServer(
    id,
    function(input, output, session) {
      observe({
        updateSelectInput(
          session,
          "variable",
          choices = names(dataset())
        )
      })
      
      reactive(input$variable)
    }
  )
}

server <- function(input, output, session) {
  dataset <- selectDatasetServer("dataset_select")
  
  variable <- selectVariableServer(
    "variable_select",
    dataset$data
  )
  
  output$module_out <- renderText({
    paste0(
      "Selected variable `", variable(), "` ",
      "from dataset `", dataset$name(), "`"
    )
  })
}
```
::::

:::::

::: aside
📝 session_05/module_reactive.R
:::

## Namespacing and dynamic UI

::::: {.columns}

:::: {.column width="40%"}
::: {.incremental}
* UI created inside the server function needs to be namespaced
* Use the `session` object to access the module namespace
  - `session$ns(id)`
:::
::::

:::: {.column width="60%"}
```{.r code-line-numbers="|5-14|8,9"}
dynamicTableServer <- function(id) {
  moduleServer(
    id,
    function(input, output, session) {
      output$custom_ui <- renderUI({
        if (input$render_ui) {
          tagList(
            selectInput(session$ns("dataset"), label = "Dataset", choices = ls("package:datasets")),
            tableOutput(session$ns("table"))
          )
        } else {
          NULL
        }
      })
      
      output$table <- renderTable({
        dataset <- get(input$dataset, "package:datasets")
        dataset
      })
    }
  )
}
```
::::

:::::

::: aside
📝 session_05/module_render_ui.R
:::

## Nesting modules

::::: {.columns}

:::: {.column width="40%"}
::: {.fragment}
Use modules inside other modules
:::

::: {.incremental}
* Modules are like functions
  - Composable
  - Multiple levels of nesting
* Modules are like apps
  - UI and server pairing
:::
::::

:::: {.column width="60%"}
```{.r code-line-numbers="|4-5|16-18"}
selectDataVarUI <- function(id) {
  ns <- NS(id)
  tagList(
    selectDatasetUI(ns("dataset_select")),
    selectVariableUI(ns("variable_select")),
    
    verbatimTextOutput(ns("module_out"), placeholder = TRUE),
    tableOutput(ns("table"))
  )
}

selectDataVarServer <- function(id) {
  moduleServer(
    id,
    function(input, output, session) {
      dataset <- selectDatasetServer("dataset_select")
      
      variable <- selectVariableServer("variable_select", dataset$data)
      
      output$module_out <- renderText({
        paste0(
          "Selected variable `", variable(), "` ",
          "from dataset `", dataset$name(), "`"
        )
      })
      
      output$table <- renderTable({
        dataset$data()
      })
      
      list(
        data = dataset$data,
        name = dataset$name,
        var = variable
      )
    }
  )
}
```
::::

:::::

::: aside
📝 session_05/module_nested.R
:::

# Data for Shiny applications

## Approaches

::::: {.columns}

:::: {.column width="50%"}
::: {.fragment fragment-index=1}
Local data files
: Simple pre-prepared data that is application specific
:::

::: {.fragment fragment-index=3}
* Data is specific to this application
  - Not shared!
:::
::: {.fragment fragment-index=4}
* Changes with the application
  - No change between dev cycles
  - Read-only
:::
::::

:::: {.column width="50%"}
::: {.fragment fragment-index=2}
External data sources
: Data that changes over time outside of the Shiny app
:::

::: {.fragment fragment-index=3}
* Shared data sources
  - Used by multiple processes
:::
::: {.fragment fragment-index=4}
* Changes over time
  - Updated outside of dev cycle
  - User actions update data
::::
:::

:::::

## Local data files

Data that *lives with the application*

::::: {.columns}

:::: {.column width="50%"}
::: {.incremental}
* Intrinsically linked to the application set up
  - Configuration
* Strictly read only
  - Re-deploying the application replaces the data
:::
::::

:::: {.column width="50%"}
::: {.incremental}
* Local files
  - rds, csv, SQLite database
* Pre-processed
  - Do as much processing as possible outside of the app
  - Minimise the work done during the user session
:::
::::

:::::

::: {.fragment}
Best practices for incorporating static files are discussed in session 6!
:::

## External data sources

Data that *changes over time*

Data that is *shared*

::::: {.columns}

:::: {.column width="50%"}
::: {.incremental}
* Changes independent of the app
  - Updated by an external system
  - Reporting
* Read write
  - Updated by users
  - CRUD applications
:::
::::

:::: {.column width="50%"}
::: {.incremental}
* External sources
  - Databases
  - Web services / APIs
* External processing
  - Databases are much more efficient than R!
:::
::::

:::::

# Working with databases

## Databases and R

Several R packages are available for accessing databases

::: {.fragment}
We prefer 📦 [DBI](https://dbi.r-dbi.org/)
:::

::: {.incremental}
* Developed and maintained by RStudio
* Consistent interface for accessing many different database platforms
* Integrated with other packages
:::

::: {.fragment}
Our focus is Shiny database usage - see the RStudio [Databases using R](https://solutions.rstudio.com/db/) documentation for a good general overview of database access
:::

## Shared connections with pool

Managing database connections is a pain!

::: {.incremental}
* Creating new connection takes time
* Sharing connections between sessions is hard
:::

&nbsp;

::: {.fragment}
📦 [pool](https://rstudio.github.io/pool/) takes care of this for you
:::

::: {.incremental}
* Creates a *pool* of database connections
* Opens and closes database connection as necessary
:::

## Using pool

::::: {.columns}

:::: {.column width="40%"}
::: {.incremental}
* Drop in replacements for DBI functions
  - `dbPool()` for `dbConnect()`
  - `poolClose()` for `dbDisconnect()`
* Use the pool as you would a DBI connection
:::
::::

:::: {.column width="60%"}

&nbsp;

```{.r code-line-numbers="|1-7|9|11"}
pool <- dbPool(
  drv = RMySQL::MySQL(),
  dbname = "shinydemo",
  host = "shiny-demo.csa7qlmguqrf.us-east-1.rds.amazonaws.com",
  username = "guest",
  password = "guest"
)

dbGetQuery(pool, "SELECT * from City LIMIT 5;")

poolClose(pool)
```
::::

:::::

## Avoiding SQL injections

User inputs can't be trusted - 🌐 [XKCD: Exploits of a Mom](https://xkcd.com/327/)

&nbsp;

::: {.fragment}
Two methods to avoid malicious user inputs with DBI
:::

::::: {.columns}

:::: {.column width="50%"}
::: {.fragment}
* Parameterised queries
  - Pass variable arguments to the database server as *parameters*
  - `dbGetQuery(params = )`
:::
::::

:::: {.column width="50%"}
::: {.fragment}
* Escaping user inputs
  - Apply appropriate escaping rules for the DBI connection
  - `sqlInterpolate()`
  - `glue::glue_sql()`
:::
::::

:::::

## Performance considerations

Working with a database has its own performance considerations

::: {.incremental}
* Data aggregation queries can be expensive
* Summarising data across multiple tables can require expensive joins
:::

::: {.fragment}
Design your database and queries to minimise processing time
:::

::::: {.columns}

:::: {.column width="50%"}
::: {.incremental}
* Pre-calculate tables
  - Pre-aggregate data
  - Pre-process expensive joins
  - Scheduled re-calculation
:::
::::

:::: {.column width="50%"}
::: {.incremental}
* Index your tables!
* Avoid expensive filters (e.g. `like`)
* Profile your queries
:::
::::

:::::

# Working with user data

## Validating user inputs

::: {.incremental}
* Check that required values have been provided
  - Don't try to produce outputs before input exists
  - Avoid Shiny printing `NULL` to output
* User inputs can't be trusted!
  - Make sure they're in the expected range
  - Provide feedback to the user
:::

## Values exist

Check that any value has been provided

::::: {.columns}

:::: {.column width="40%"}
::: {.incremental}
* `req()` - Require a non-empty value for an input before processing
* Stops expression
  - Clear output by default
  - `cancelOutput = TRUE` to leave existing
:::
::::

:::: {.column width="60%"}
```{.r code-line-numbers="|13-26|14-15"}
server <- function(input, output, session) {
  # Return the selected dataset
  dataset <- reactive({
    dataset <- as.data.frame(get(input$dataset, "package:datasets"))
  })
  
  # Render selected dataset
  output$table <- renderTable({
    dataset()
  })
  
  # Render a summary of the selected variable
  output$summary <- renderPrint({
    # Don't run if a dataset hasn't been selected
    req(dataset())
    
    # Return an error message if the selected variable isn't in dataset()
    validate(
      need(
        input$variable %in% names(dataset()),
        "Select a valid variable"
      )
    )
    
    summary(dataset()[input$variable])
  })
}
```
::::

:::::

::: aside
📝 session_05/user_data_req_validate.R
:::

## Values are valid

Check that valid values have been provided

::::: {.columns}

:::: {.column width="40%"}
::: {.incremental}
* `validate()`
  - Function producing message on failure
  - Printed to output area
* `need()`
  - Helper function
  - Condition and message
:::
::::

:::: {.column width="60%"}
```{.r code-line-numbers="|13-26|17-23"}
server <- function(input, output, session) {
  # Return the selected dataset
  dataset <- reactive({
    dataset <- as.data.frame(get(input$dataset, "package:datasets"))
  })
  
  # Render selected dataset
  output$table <- renderTable({
    dataset()
  })
  
  # Render a summary of the selected variable
  output$summary <- renderPrint({
    # Don't run if a dataset hasn't been selected
    req(dataset())
    
    # Return an error message if the selected variable isn't in dataset()
    validate(
      need(
        input$variable %in% names(dataset()),
        "Select a valid variable"
      )
    )
    
    summary(dataset()[input$variable])
  })
}
```
::::

:::::

::: aside
📝 session_05/user_data_req_validate.R
:::

## Input feedback

📦 [shinyFeedback](https://merlinoa.github.io/shinyFeedback/) - styled user feedback on UI controls

::::: {.columns}

:::: {.column width="40%"}
::: {.incremental}
* `useShinyFeedback()` in UI to enable
* `showFeedback*()` to add feedback to control
* `hideFeedback()` to remove
:::
::::

:::: {.column width="60%"}
```{.r code-line-numbers="|4|14-24|16-19|21"}
library(shinyFeedback)

ui <- fluidPage(
  useShinyFeedback(),
  
  numericInput(
    "positive_number",
    "Enter a number",
    value = -5
  )
)

server <- function(input, output, session) {
  observe({
    if (input$positive_number < 0) {
      showFeedbackWarning(
        "positive_number",
        "Please enter a positive number"
      )
    } else {
      hideFeedback("positive_number")
    }
  }) |>
    bindEvent(input$positive_number)
}
```
::::

:::::

::: aside
📝 session_05/user_data_shiny_feedback.R
:::

## User notifications

Use Shiny's default notification system

::::: {.columns}

:::: {.column width="40%"}
::: {.incremental}
* `showNotification()`
  - Styled notifications
  - Temporary or permanent
  - Closable
* `removeNotification()`
  - Remove an existing notification by ID
:::
::::

:::: {.column width="60%"}
::: {.fragment}
```{.r code-line-numbers=false}
showNotification(
  ui,
  action = NULL,
  duration = 5,
  closeButton = TRUE,
  id = NULL,
  type = c("default", "message", "warning", "error"),
  session = getDefaultReactiveDomain()
)
```
:::

&nbsp;

::: {.fragment}
```{.r code-line-numbers=false}
removeNotification(
  id,
  session = getDefaultReactiveDomain()
)
```
:::
::::

:::::

::: aside
📝 session_05/user_notifications.R
:::

# Wrapping up

## Homework - Readings

### Mastering Shiny

**[Chapter 19: Shiny modules](https://mastering-shiny.org/scaling-modules.html)**

**[Chapter 5: User feedback](https://mastering-shiny.org/action-feedback.html)**

### Shiny articles

**[Modularizing Shiny app code](https://shiny.rstudio.com/articles/modules.html)**

**[Communication between modules](https://shiny.rstudio.com/articles/communicate-bet-modules.html)**

**[Data](https://shiny.rstudio.com/articles/#data)**

**[RStudio: Databases using R](https://solutions.rstudio.com/db/)**

## Homework - Exercise

Continue with your shiny application from session 4

::::: {.columns}

:::: {.column width="50%"}
::: {.fragment}
* Use a Shiny module
  - Convert some of your existing code to modules
  - Try creating modules with inputs and outputs
:::
::::

:::: {.column width="50%"}
::: {.fragment}
* Validate your inputs
  - Use `req()` and `validate()` to check for valid data
  - Provide user feedback for invalid inputs with shinyFeedback
:::
::::

:::::

## Goodbye!

&nbsp;

### Next Session

6. *Application structure and workflow*
